---
title: "The Success of College Basketball Teams can be predicted using several advanced team statistics"
subtitle: "A Linear Model Approach"
author: 
  - Thomas D'Onofrio
thanks: "Code and data are available at: https://github.com/TDonofrio62/Paper-On-NCAAB-Stats."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(palmerpenguins)
```

# Introduction

March is a very exciting time for sports fans in the US and many other countries, as it is when the NCAA Men's Division 1 Basketball Tournament, more commonly known as March Madness, is played. 68 college basketball teams go head to head, playing single elimination games until there is only one team remaining, making for some of the most intense and unpredictable sports games in the world. As a result of this, March Madness draws in a lot of viewers. 16.9 million people tuned in to the 2021 March Madness finals, making it the 2nd most viewed North American sports event only behind the Superbowl [@citeBaker]. A big reason for the massive following is that many people create brackets for the tournament, trying to predict the results. With 9.2 quintillion different bracket possibilities, all the insane upsets that nobody sees coming, and the overall chaotic nature of the tournament, creating a bracket and seeing how well you do is very appealing to both sports fans and regular people, causing all the hype around March Madness [@citeGeiling]. People put a lot of work into their brackets, comparing teams to one another and trying to figure out who has the best odds of winning. With this in mind, is there a way fans can more successfully predict which teams will have more success than others in the tournament?

In this paper, we attempt to answer this intriguing question by analyzing a variety datasets from the college basketball section of Sports Reference [@citeSR], using R [@citeR]. Datasets containing college basketball statistics from 2017 to 2020 were used, where each dataset contains many different advanced team statistics for each NCAA Division 1 school with a Men's basketball team in a specific season. The data was first graphically analyzed to determine which team statistics have a strong or noticeable relationship with how many games a team wins. These statistics were then used to create a linear model that attempts to predict winning percentage, the response variable that is being used to measure team success, where only the best and most necessary team statistics were included. As a result of this, it was determined that the following statistics [Insert statistics here once model is completed], play the most important factor in leading to team success. Thus, when filling out their brackets, fans can compare these statistics across teams to determine which teams have a good chance of going far or even winning the tournament, as opposed to just looking at wins or their seeding like many people do.

Unfortunately, there are some implications with these findings. 

The rest of this paper will be as follows: Section \@ref(data) goes over the datasets being used in the paper, and analyzes the data with the use of tables and graphs to see which team statistics have a relationship with winning percentage. Section \@ref(model) includes the creation of a linear regression model that uses only the most important team statistics as explanatory variables to predict winning percentage, the response variable that determines team success. Section \@ref(results) interprets the coefficients and provides a statistical analysis of the model in order to understand what is being done. Finally, Section \@ref(discussion) discusses what exactly the model and its outcome is telling us, what can be learned from it, and what implications come along with the findings.

# Data

Our data is of penguins (Figure \@ref(fig:bills)).

```{r bills, fig.cap="Bills of penguins", echo = FALSE}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = FALSE) +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

Also bills and their average (Figure \@ref(fig:billssssss)). (Notice how you can change the height and width so they don't take the whole page?)

```{r billssssss, fig.cap="More bills of penguins", echo = FALSE, fig.width=8, fig.height=4}

# This needs to be about the random data tha tI sasmpled?
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = FALSE) +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk way more about it. 



# Model

\begin{equation}
Pr(\theta | y) = \frac{Pr(y | \theta) Pr(\theta)}{Pr(y)}  (\#eq:bayes)
\end{equation}

Equation \@ref(eq:bayes) seems useful, eh?

Here's a dumb example of how to use some references: In paper we run our analysis in `R` [@citeR]. We also use the `tidyverse` which was written by @thereferencecanbewhatever If we were interested in baseball data then @citeLahman could be useful. 

We can use maths by including latex between dollar signs, for instance $\theta$.


# Results

# Discussion

## First discussion point

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional details


\newpage


# References


